Model complexity and model choice for animal movement models

### 2016-06-08 guelph_talk/guelph_panther.Rnw
... more ...

<<setup,echo=FALSE,message=FALSE>>=
library(knitr)
opts_chunk$set(echo=FALSE,error=FALSE,
               fig.align="center",out.width="0.8\\textwidth",
               fig.width=6,fig.height=4)
library(deSolve)
library(reshape)
library(lattice)
library(latticeExtra)
library(emdbook)
library(nlme)
library(bbmle)
library(tikzDevice)
## http://yihui.name/en/2011/04/produce-authentic-math-formulas-in-r-graphics/
options(tikzMetricPackages = c( "\\usepackage[utf8]{inputenc}",
                "\\usepackage[T1]{fontenc}", "\\usetikzlibrary{calc}",
                "\\usepackage{amssymb}"))

@@

\begin{frame}
\frametitle{Acknowledgements}
\begin{description}
\item[People] \textbf{Michael Li}, \textbf{Madelon van de Kerk}, Dave Onorato, Madan Oli
\item[Agencies] US Fish and Wildlife, US Geological Survey, US National Park Service
\item[Funding] NSERC Discovery grant, NSF IGERT program
\end{description}

@@

\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item observations:  \\
e.g. mass mark-recapture, longitudinal density,
direct observation, telemetry (VHF, GPS)
\item most methods provide a sequence of times and locations for each individual
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
(sequence pix)
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item summaries:
\begin{itemize}
\item home range \\
(convex hull, kernel density estimate, etc.)
\item root-mean-squared displacement
\item step length and turning angle
\end{itemize}
\item covariates: \\
e.g. habitat map,  \\
individual characteristics (sex, age, weight \ldots)
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
(step length/turning angle pix)
\end{column}
\end{columns}
\end{frame}

@@

\begin{columns}
  \begin{column}{0.5\textwidth}
    \begin{itemize}
    \item panthers tracked, captured
    \item GPS collars
    \item 18 males (13 male, 5 female, 1-15 years old)
    \item 3200 panther days, hourly/bihourly; 49000 locations
    \item ?? per panther
    \end{itemize}
  \end{column}
  \begin{column}{0.5\textwidth}
    \includegraphics[trim={10cm 0 10cm 0},clip,width=\textwidth]{pix/panther_collar.jpg}
  \end{column}
\end{columns}
\end{frame}

@@

\includegraphics[trim={0 6.5cm 0 0}, clip, width=\textwidth]{pix/jae_fig2.png}

@@

\includegraphics[width=\textwidth]{pix/jae_fig3.png}

@@

\begin{frame}
\frametitle{Transition parameters}

picture/table here of transition parameters (network diagram??)
\end{frame}

@@

\begin{frame}
\frametitle{Model complexity}

@@

\begin{frame}
\frametitle{Broadening the model}

Attempting to fix these problems:
\begin{itemize}
\item extend the model to allow covariates
\item specifically, allow for diurnal variation
\begin{itemize}
\item simplify model (log-Normal step length only)
\item \emph{fixed} state-specific emissions parameters\\
(step length mean and std dev)
\item time-varying transition parameters
\item also try \emph{finite mixture models} \\
(independent occupancy)
\end{itemize}
\item how much does this help?
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Temporal models}

figure showing alternative temporal models
\end{frame}

@@

\begin{frame}
\frametitle{Model complexity}
Figure showing BIC plots for all cats tried
\end{frame}

\begin{frame}
\frametitle{Model complexity}
Simulation results
\end{frame}

@@

\begin{frame}
\frametitle{Animal movement: open challenges}
\begin{itemize}
\item Cognition
\item Intraspecific interaction/collective movement \citepx{delgado_statistical_2014}
\item Continuous-time movement models \citepx{calabrese_ctmm_2016}
\item Edges, barriers, and corridors \citepx{beyer_you_2016}
\item Efficient (big-data) approaches \citepx{brillinger_three_2008}
\item Putting it all together ...
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item simple model families +  \\
  model misspecification $\to$ \\
  overparameterization
\item protocols and tools for model checking \citepx{potts_generalized_2014};
score tests?
\item flexible computational frameworks
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{an aside on AIC vs BIC}

@@

@@

\begin{itemize}
\item cross-validation \citepx{wenger_assessing_2012}
\item diagnostic plots
\item score tests?
\end{itemize}

>>>>

<<setup,echo=FALSE,message=FALSE>>=
library(knitr)
opts_chunk$set(echo=FALSE,error=FALSE,
               fig.align="center",out.width="0.8\\textwidth",
               fig.width=6,fig.height=4)
library(deSolve)
library(reshape)
library(lattice)
library(latticeExtra)
library(emdbook)
library(nlme)
library(bbmle)
library(tikzDevice)
library(plotrix)
## http://yihui.name/en/2011/04/produce-authentic-math-formulas-in-r-graphics/
options(tikzMetricPackages = c( "\\usepackage[utf8]{inputenc}",
                "\\usepackage[T1]{fontenc}", "\\usetikzlibrary{calc}",
                "\\usepackage{amssymb}"))

@@

\begin{frame}
\frametitle{Acknowledgements}
\begin{description}
\item[People] \textbf{Michael Li}, \textbf{Madelon van de Kerk}, \\
Dave Onorato, Madan Oli
\item[Agencies] US Fish and Wildlife Service, US Geological Survey, US National Park Service
\item[Funding] NSERC Discovery grant, NSF IGERT program
\end{description}

@@

\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item observations:  \\
e.g. mass mark-recapture, longitudinal density,
direct observation, telemetry (VHF, GPS)
\item most methods provide a sequence of times and locations for each individual
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
<<arcpic,echo=FALSE,fig.width=3,fig.height=3>>=
x <- c(0,-0.5,0.5,0)
y <- c(0,0.5,1,1.5)
par(mar=c(1,1,1,1),cex=1.5,xpd=NA)
plot(x,y,axes=FALSE,ann=FALSE,pch=16)
eps <- 0.04
arrows(x[-4]+c(-1,1,-1)*eps,y[-4]+eps,x[-1]+c(1,-1,1)*eps,y[-1]-eps,lwd=2)
draw.arc(x[2],y[2],radius=0.2,angle1=-pi/6,angle2=pi/10,col=4)
text(x[2]+0.28,y[2],expression(theta[1]),col=4)
draw.arc(x[3],y[3],radius=0.2,angle1=pi-pi/6,angle2=pi+pi/8,col=4)
text(x[3]-0.28,y[3],expression(theta[2]),col=4)
text((x[2]+x[3])/2,(y[2]+y[3])/2,pos=3,expression(s[1]),col=2)
text((x[3]+x[4])/2,(y[3]+y[4])/2,pos=3,expression(s[2]),col=2)
posvec <- c(4,2,4,3)
for(i in 1:4)
    text(x[i], y[i], substitute(t[j],
                                 list(j = i)),pos=posvec[i])
@ 
\end{column}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{}
\begin{columns}
\begin{column}{0.5\textwidth}
\begin{itemize}
\item summaries:
\begin{itemize}
\item home range \\
(convex hull, kernel density estimate, etc.)
\item root-mean-squared displacement
\item step length and turning angle
\end{itemize}
\item covariates: \\
e.g. habitat map,  \\
individual characteristics (sex, age, weight \ldots)
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth}
<<getdat,echo=FALSE,cache=TRUE,warning=FALSE>>=
load("../allcats.RData")  ## all data
d1 <- subset(dat,cat==1)
g1 <- ggplot(d1,aes(x=LogDist))+
    geom_histogram(binwidth=0.1)+
    labs(x=expression("step length"~(log[10]*m)))
g2 <- ggplot(d1,aes(x=Turningangle))+
    geom_histogram(binwidth=pi/16,alpha=0.2,colour="black",lwd=0.2)+
    coord_polar()+labs(x="",y="")+
    scale_x_continuous(breaks=c(-1,-0.5,0,0.5)*pi,
                       labels=c(180,270,0,90))
print(g1)
@

<<g2plot,warning=FALSE>>=
print(g2)
@ 
\end{column}
\end{columns}
\end{frame}

@@

\begin{columns}
  \begin{column}{0.5\textwidth}
    \begin{itemize}
    \item panthers tracked, captured
    \item GPS collars
    \item 18 males (13 male, 5 female, 1-15 years old)
    \item 3200 panther days, hourly/bihourly; 49000 locations
    \item ?? per panther
    \end{itemize}
  \end{column}
  \begin{column}{0.5\textwidth}
    \includegraphics[trim={10cm 5cm 10cm 0},clip,width=\textwidth]{pix/panther_collar.jpg}
  \end{column}
\end{columns}
\end{frame}

@@

\includegraphics[width=\textwidth]{pix/distpars.pdf}

@@

\includegraphics[width=\textwidth]{pix/parplot.pdf}

@@

%% \begin{frame}
%% \frametitle{Transition parameters}

%% picture/table here of transition parameters (network diagram??)
%% \end{frame}

@@

\begin{frame}
\frametitle{Model complexity (bad news)}

@@

\begin{frame}
\frametitle{Expanding the model}

Attempting to fix these problems:
\begin{itemize}
\item extend the model to allow covariates
\pause
\item specifically, allow for diurnal variation
\begin{itemize}
\item simplify model (log-Normal step length only)
\item \emph{fixed} state-specific emissions parameters\\
(step length mean and std dev)
\item time-varying transition parameters
\item also try \emph{finite mixture models} \\
(independent occupancy)
\end{itemize}
\item how much does this help?
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Temporal models}
<<tempmodel>>=
bpar <- c(1,0.2,0.8,0.3)
bseg <- c(10,15,20,24)
hourpars <- c(0,  0.5, 0.8, 0.9, 0.87, 0.8,
              0.3,0.4, 0.3, 0.5, 0.4,  0.7,
              0.9,0.8, 0.7, 0.6, 0.7,  0.6, 0.7,
              0.5,0.1,0.3,0.4,0.5)
dd <- data.frame(time=seq(0,to=23.9,length=101)) %>%
    mutate(sin=((1+sin(2*pi*time/24))/2),
           quad=1-(time/24)^2,
           block=ifelse(time<bseg[1],bpar[1],
                   ifelse(time<bseg[2],bpar[2],
                      ifelse(time<bseg[3],bpar[3],bpar[4]))),
           hourly=hourpars[floor(time)+1]) %>%
    gather(model,val,-time)
ggplot(dd,aes(time,val,colour=model))+geom_line(size=1.5)+
    scale_x_continuous(limits=c(0,24),breaks=seq(0,24,by=6))+
    theme(axis.ticks.y=element_blank(),
          axis.text.y=element_blank())+
    labs(y="")
@ 
\end{frame}

@@

\begin{frame}
\frametitle{Diurnal model: conclusions}
\begin{itemize}
\item diurnal structure greatly improves fit ($\Delta \textrm{BIC} \approx 500$)
\item slightly improves latent-state issue ($n=6 \to 5$)
\item lots left to do!
\begin{itemize}
\item seasonal variation
\item incorporate habitat, home range behaviour
\item etc. etc. etc.
\end{itemize}
\end{itemize}
\end{frame}

@@

@@

\begin{itemize}
\item simple model families +  \\
  model misspecification $\to$ \\
  overparameterization
\item Gelman: ``Sample sizes are never large'':
(\href{http://andrewgelman.com/2005/07/31/n_is_never_larg/}{blog post})
\begin{quote}
$N$ is never enough because if it were ``enough'' you'd already be on to the next problem for which you need more data.
\end{quote}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{An aside on AIC vs BIC}

@@

\begin{frame}
\frametitle{Animal movement: open challenges}
\begin{itemize}
\item Cognition/memory \citepx{bracis_memory_2015}
\item Intraspecific interaction/collective movement \citepx{delgado_statistical_2014}
\item Continuous-time movement models \citepx{calabrese_ctmm_2016}
\item Edges, barriers, and corridors \citepx{beyer_you_2016}
\item Efficient (big-data) approaches \citepx{brillinger_three_2008}
\item Putting it all together ...
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item cross-validation \citepx{wenger_assessing_2012}
\item protocols and tools for model checking \citepx{potts_generalized_2014};
score tests?
\item flexible computational frameworks \\
(ecologists can't afford consultants/ \\
there are too many species out there)
\end{itemize}
